---
title: "Regression Analysis Calculation"
fontsize: 11pt
geometry: a4paper
output: pdf_document
---

# Introduction

This document calculates and explains the following:
- **Variance** (\(Var(X)\)): Measures how much X values deviate from their mean.
- **Covariance** (\(Cov(X, Y)\)): Measures how X and Y vary together.
- **Regression slope** (\(\beta_1\)): The rate at which Y changes with X.

The dataset consists of High School GPA (\(X\)) and First Year GPA (\(Y\)) for 10 students.

# Dataset

```{r}
students <- data.frame(
  Student = paste("Student", 1:10),
  High_School_GPA = c(3.0, 3.2, 3.5, 3.8, 3.6, 3.1, 3.9, 3.4, 3.7, 3.2),
  First_Year_GPA = c(2.8, 3.0, 3.2, 3.8, 3.6, 3.3, 3.9, 3.7, 4.0, 2.5)
)

knitr::kable(students, caption = "High School GPA and First Year GPA Dataset")
```

---

# Step 1: Variance of X

Variance measures how much the values of X (High School GPA) deviate from their mean.

## Formula
\[
\text{Var}(X) = \frac{1}{n} \sum_{i=1}^{n} (X_i - \bar{X})^2
\]

## Steps
1. Compute the mean of X (\( \bar{X} \)).
2. Compute deviations from the mean and square them.
3. Compute the average of the squared deviations.

```{r}
# Variance calculation
X <- students$High_School_GPA
mean_X <- mean(X)
students$Deviation_X <- paste(X, "-", round(mean_X, 2), "=", round(X - mean_X, 2))
students$Squared_Deviation_X <- (X - mean_X)^2
variance_X <- mean((X - mean_X)^2)

knitr::kable(students[, c("Student", "High_School_GPA", "Deviation_X", "Squared_Deviation_X")],
             caption = "Step-by-step Variance Calculation for High School GPA")
```

The variance of X (\( \text{Var}(X) \)) is:

```{r}
variance_X
```

# Step 2: Covariance of X and Y

Covariance measures how X and Y vary together. A positive covariance indicates that when X increases,Y tends to increase as well.

## Formula
\[
\text{Cov}(X, Y) = \frac{1}{n} \sum_{i=1}^{n} (X_i - \bar{X})(Y_i - \bar{Y})
\]

## Steps
1. Compute the means of X and Y (\( \bar{X} \) and \( \bar{Y} \)).
2. Compute deviations from the mean for both variables.
3. Compute the product of deviations for each pair.
4. Compute the average of these products.

```{r}
# Covariance calculation
Y <- students$First_Year_GPA
mean_Y <- mean(Y)
students$Deviation_Y <- paste(Y, "-", round(mean_Y, 2), "=", round(Y - mean_Y, 2))
students$Product_Deviation <- (X - mean_X) * (Y - mean_Y)
covariance_XY <- mean((X - mean_X) * (Y - mean_Y))

knitr::kable(students[, c("Student", "High_School_GPA", "First_Year_GPA", "Deviation_X", "Deviation_Y", "Product_Deviation")],
             caption = "Step-by-step Covariance Calculation between High School GPA and First Year GPA")
```

The covariance of X and Y (\( \text{Cov}(X, Y) \)) is:

```{r}
covariance_XY
```

---

# Step 3: Regression Slope

The regression slope \( \beta_1 \) is calculated as:
\[
\beta_1 = \frac{\text{Cov}(X, Y)}{\text{Var}(X)}
\]

It represents the rate at which Y changes for a one-unit increase in X.

```{r}
slope <- covariance_XY / variance_X
slope
```
