---
title: Horizontal Stratification of Higher Education and Gender Earnings Gap in the
  Early Labor Market
fontsize: 11pt
geometry: a4paper, twoside, left=2.5cm, right=2.5cm, top=2cm, bottom=2.8cm, headsep=1.35cm,
  footskip=1.6cm
output: pdf_document
header-includes:
- \usepackage{titling}
- \usepackage{amsmath}
- \usepackage{amssymb}
---

\noindent\makebox[\textwidth][c]{
  \begin{minipage}[t]{0.8\textwidth}
    \centering
    \vspace{-6em}
    \Large{Inchan Hwang, \texttt{inchan@yonsei.ac.kr}}
  \end{minipage}
}

\begin{center}
\vspace{-6em}
\large{2 Jan 2025}
\end{center}

# Basics of Statistics

Statistics is a tool used to extract information from data and support decision-making. Statistics are widely used in the social sciences to analyze patterns, understand relationships, and draw conclusions from data. Fundamentally, traditional statistics aim to summarize data effectively.
This session is introduction of two basic statistical methods: (1) how to compare several group's outcome variables (t-test and ANOVA) (2) how to examine the relationship between explanatory variable and outcome variable. (regression and logistic regression)

For instance, consider a class of fifteen students with their scores in Korean and Math. These scores can be grouped by class (Class 1, Class 2, and Class 3) and analyzed to better understand the overall performance of each group.

In this section, we will explore how data can be summarized using these basic statistical measures.

- **Population**: The entire set of data we are interested in.
- **Sample**: A subset of the population selected for statistical analysis.
- **Mean**: A measure representing the central value of the data.
- **Standard Deviation**: A measure indicating how much the data varies from the mean.

```{r, echo=TRUE}
# Generating example data
data <- data.frame(
  Student = paste("Student", 1:15), # Names of students
  Class = rep(c("Class 1", "Class 2", "Class 3"), each = 5), # Class assignments
  Korean = c(100, 100, 100, 100, 95, 82, 87, 75, 89, 91, 80, 50, 66, 77, 84), # Korean scores
  Math = c(100, 100, 100, 90, 95, 84, 79, 90, 86, 83, 78, 82, 88, 81, 85)    # Math scores
)

# Viewing the data
data

# Calculating mean and standard deviation for each subject
mean(data$Korean)
sd(data$Korean)
mean(data$Math)
sd(data$Math)
```
Summarizing data using **mean** and **standard deviation** allows us to capture the central tendency and variability within each class for both Korean and Math scores. This provides a clear picture of how students in Class 1, Class 2, and Class 3 perform overall.

# t-Test

## What is a t-Test?

t-Test is a statistical method used to compare the means of two groups. It helps to determine whether there is a statistically significant difference between the groups. The formula for the independent t-test is:

\[
 t = \frac{\bar{X}_1 - \bar{X}_2}{\sqrt{\frac{s_1^2}{n_1} + \frac{s_2^2}{n_2}}}
\]

Where:
- \( \bar{X}_1 \) and \( \bar{X}_2 \): Means of the two groups
- \( s_1^2 \) and \( s_2^2 \): Variances of the two groups
- \( n_1 \) and \( n_2 \): Sample sizes of the two groups

### Applying t-Test to the Data

Now, we will apply the t-test to compare the Math scores between Class 1 and Class 2:

```{r, echo=TRUE}
# Independent t-test comparing Math scores between Class 1 and Class 2
independent_ttest <- t.test(Math ~ Class, data = subset(data, Class %in% c("Class 1", "Class 2")), var.equal = TRUE) # Testing mean difference between classes
independent_ttest # Display the test result
```

The results of the t-test show whether there is a significant difference in the Math scores between the two classes.


# Analysis of Variance (ANOVA)

## What is ANOVA?

ANOVA (Analysis of Variance) is used to compare the means of three or more groups. It tests whether there are significant differences among group means. The formula for ANOVA can be expanded as:

\[
F = \frac{\text{Between-group variability}}{\text{Within-group variability}} = \frac{\frac{SS_{between}}{df_{between}}}{\frac{SS_{within}}{df_{within}}} = \frac{MS_{between}}{MS_{within}}
\]

Where:
- \(SS_{between}\): Sum of squares between groups (measures variability due to group differences)
- \(df_{between}\): Degrees of freedom for the between-group variability \((k - 1)\), where \(k\) is the number of groups
- \(SS_{within}\): Sum of squares within groups (measures variability within each group)
- \(df_{within}\): Degrees of freedom for the within-group variability \((N - k)\), where \(N\) is the total number of observations

### One-Way ANOVA

Compares the means of groups based on one factor.

```{r, echo=TRUE}
# Performing ANOVA on the example data
anova_result <- aov(Math ~ Class, data = data) # Testing mean differences in Math scores between classes
summary(anova_result) # Display the ANOVA table
```

# Regression

## What is Regression?
